Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-13 manuscript/manuscript.tex
Fix some LaTeX styling

\title[Variable Density Tesseroids]{
    Gravitational field calculation in spherical coordinates using variable
    densities in depth
}
\author[S.R. Soler, A. Pesce, L. Uieda and M.E. Gimenez]{
    Santiago R. Soler$^{1,2}$, Agustina Pesce$^{1,2}$, Leonardo Uieda$^3$ and
    Mario E. Gimenez$^{1,2}$ \\
    $^1$CONICET, Argentina. e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de
    San Juan, Argentina\\
    $^3$Department of Earth Sciences, SOEST, University of Hawai‘i at
    M\={a}noa, Honolulu, Hawaii, USA
}

@@

The lithosphere's density variation with depth has been studied for close to a
century.
Over this time period,
several density-depth relations have been proposed for different rock types
\citep[e.g.,][]{Maxant1980, Rao1986, Rao1993, Rao1994}.
Furthermore, depth-variable densities have been used in forward and
inverse gravity modelling, mostly applied to sedimentary basins
\citep{Cordell1973, Rao1986, Cowie1990, Rao1993, Rao1994, Zhang2001,
Welford2010}.
These forward gravity models have been developed for two or three dimensional
bodies in Cartesian coordinates, which limits the applications to local scales.
The advent of satellite gravimetry has provided gravity field
measurements with global coverage, enabling modelling and interpretation on regional and
global scales.
Hence, designing forward modelling methods that reproduce the gravity anomalies for
such scales is of high importance.

@@

Two recent papers present other approaches for homogeneous tesseroids gravitational
fields computation and incorporate methodologies for variable density tesseroid with
specific density profiles.
\citet{Fukushima2018} analytically integrates the volumetric integral for the
gravitational potential in the radial direction, obtaining a surface integral, which is
then numerically solved by applying a conditionally splitting and the double exponential
quadrature rule. The gradient of the potential and the Marussi tensor components are
then computed by finite differences. \citet{Fukushima2018} also generalizes his method
to an arbitrary polynomial density profile on the radial coordinate.
\citet{Lin2018} made a comparison of the different methodologies for homogeneous
tesseroid gravitational fields computation by applying different discretizations
algorithms to each one of them.
From this analysis they develop a combined method:
for computation points near the tesseroid they apply a GLQ numerical integration with
a two dimensional adaptive discretization algorithm, i.e. only in the horizontal
dimension. But, if the computation point is farther than a certain truncation distance,
then a second order Taylor series approximation is applied along with the regular
subdivision developed by \citet{Grombein2013}.
They also introduce a variation of their combined method to compute the gravitational
fields generated by tesseroids with linear density on the radial coordinate.

@@

Applying a $N$th order GLQ, we can approximate each integral in equations
\ref{eq:tesseroid-pot} and \ref{eq:tesseroid-grav} by a weighted sum of the integration
kernel evaluated on the roots of an $N$th order Legendre polynomial
\citep[p.~390]{Hildebrand1987}.
Unlike the homogeneous density case, the radial density function $\rho(r)$ must also be
included in the integration and evaluated on the Legendre polynomial roots (i.e.,
quadrature nodes).

@@

\begin{equation}
    \cos\psi_t =
        \sin\phi\sin\phi_t + \cos\phi\cos\phi_t\cos(\lambda - \lambda_t) .
\end{equation}

\textit{Step 2}:
If none of the dimensions of the tesseroid fail inequality
\ref{eq:condition}, then compute the gravitational effect of the tesseroid using a
second-order GLQ (Eq. \ref{eq:glq-var-dens}).
Add the computed effect to a running total.

\textit{Step 3}:
If any dimension fails inequality \ref{eq:condition}, split the tesseroid in half along
the offending dimensions.
Repeat steps 1-3 for all smaller tesseroids until none are left.

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]
    {figures/density-based-discretization-algorithm.pdf}
\caption{
    Example application of the density-based discretization algorithm to a non-linear
    density function.
    (a) the normalised density function $\rho_n(r')$ (blue), current boundaries of the
    tesseroid (orange dots), and the linear density function $\rho_l(r')$ (orange line).
    The dashed red line represents the maximum density difference $\Delta \rho (r')$ at
    which the tesseroid would be divided (assuming that the inequality
    \ref{eq:delta-density} is not satisfied).
    (b) second iteration of the algorithm with a new linear density function and maximum
    density difference. The tesseroid would be divided at the depth indicated by the
    dashed red line.
    (c) third iteration of the algorithm.
    (d) final output of the density-based discretization, assuming that all four new
    tesseroids satisfy inequality \ref{eq:delta-density}.
}
\label{fig:density-discretization-algorithm}
\end{figure*}

@@

The algorithm is comprised of the following steps
(Fig. \ref{fig:density-discretization-algorithm}):

@@

\textit{Step 5}:
If inequality \ref{eq:delta-density} is not satisfied, then the tesseroid is split in
two parts at the radius $r_\text{max}$ at which the maximum absolute difference (Eq.
\ref{eq:density-abs-diff}) takes place.

@@

Once all smaller tesseroids satisfy Eq. \ref{eq:delta-density},
each one is subjected to the two dimensional adaptive
discretization algorithm described earlier to calculate their gravitational effects.

@@

\citet{Uieda2016} compared the numerical integration of homogeneous density tesseroids
with the analytical solution of a spherical shell \citep{Mikuska2006, Grombein2013} in
order to obtain default values for the distance-size ratio $D$.
We will follow this idea but for our needs the spherical shell must
have the same density function of radius as our tesseroid model.
\citet{Lin2018} show the analytical solution of the gravitational potential generated by
a spherical shell with linear density in the radial coordinate.
We derive an expression for such gravitational potential in case the spherical shell has
an arbitrary density function in depth (Eq. \ref{eq:shell-pot}), and obtained the
expressions in case of linear and exponential density (see Appendix \ref{sec:shell}).

@@

\noindent
has an analytical solution given by Eq. \ref{eq:shell-pot-linear}.

The absolute density difference defined on equation
\ref{eq:density-abs-diff} will always be zero for the linear density case.
As a result, the inequality \ref{eq:delta-density} will always be satisfied and no
divisions will ever be performed.
Therefore, the two dimensional adaptive discretization algorithm is the
only mechanism that controls the accuracy of the numerical integration.
For this reason, we will only determine the minimum value of $D$ needed in order to
guarantee an acceptable accuracy while ignoring the value of $\delta$.

In order to compare the numerical results with the analytical solution we
must build spherical shell models made of tesseroids.
These shell models are determined by their inner and outer radii and by the collection
of tesseroids used to build the shell.
For this analysis we have chosen to set the outer radii of every shell model equal to
the mean Earth radius (6378.137 km), while the inner radius is determined by the desired
thickness of the shell.
We choose  different values of shell thickness (100m, 1km, 10km, 100km and 1000km) in
order to account for different computation scenarios.
Furthermore, we propose different tesseroids collections to build the shell model by
regularly splitting it along the longitudinal and latitudinal directions.
Each collection, described in Table~\ref{tab:tess-collections}, is composed by
a different number of tesseroids, i.e. with tesseroids of different sizes.
Finally, for each collection we build a shell model with the thicknesses
previously described, giving a total of 20 different shell models.

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/linear-density-diffs.pdf}
\caption{
    Differences between the gravitational fields generated by each tesseroid shell model
    and the analytical solution.
    Each model has a linear density function (Eq. \ref{eq:density-linear}).
    The computations were performed on the four grids described in
    Table~\ref{tab:grids} and using 20 shell models built with the tesseroids
    collections detailed in Table~\ref{tab:tess-collections} and five different
    thicknesses.
    Each line represents the maximum absolute difference between every shell model with
    a certain thickness.
    Due to the linearity of the density function, the density-based discretization
    algorithm is not applied.
    Differences are reported as a percentage of the shell values.
    The horizontal dashed black line represents a target difference of 0.1\%.
}
\label{fig:D-linear}
\end{figure*}

@@

We observe from Fig. \ref{fig:D-linear} that the relative errors for the
potential and the $g_z$ component fall below the 0.1\% threshold at
$D=1$ and $D=2.5$, respectively, for every shell model and computation grid.
These results also show that there's no meaningful dependency between the accuracy of
the method and the thickness of the shell, the collection of tesseroids used to build
it or the height of the computation grid (the results for the Satellite grid and the
Global grid are very similar).

@@

We compute the relative difference between the numerical and analytical results for the
gravitational potential ($V$) and the vertical component of its gradient ($g_z$) on the
computation grids detailed in Table~\ref{tab:grids} and using shell models with the same
thicknesses as the linear density case.
Because this is a time consuming computation, we limit the analysis only to the second
collection of tesseroids in Table~\ref{tab:tess-collections}.
For the sake of brevity, we compute the maximum differences between the results
involving each computation grid, shell thickness and value of the $b$ factor.
The results are shown in Fig.~\ref{fig:grid-search}, in which points inside the dotted
lines are the ones that present a numerical error lower than the 0.1\% threshold.
Also shown in Fig. \ref{fig:grid-search} are the $D$ values determined for the linear
density function in the previous section ($D_\text{linear}$).
The values of $D_\text{linear}$ are within the 0.1\% threshold and are the optimum
values of $D$ for $V$ and $g_z$.
These results indicate that the values of $D_\text{linear}$ can be safely extrapolated
to the exponential case.

@@

\begin{figure}
\centering
\iftwocol{
\includegraphics[width=0.9\linewidth]
    {figures/grid-search.pdf}
}{
\includegraphics[width=0.5\linewidth]
    {figures/grid-search.pdf}
}
\caption{
    Numerical error exploration in the $D$-$\delta$ space.
    The percentage difference values were obtained from the comparison
    between the analytical solution and the numerical approximation of
    the gravitational fields ($V$, $g_z$ and $g_{zz}$) generated by a
    spherical shell with an exponential density function (Eq. \ref{eq:density-exp}).
    These comparisons were carried out on the ``Big Grid'' (Table
    \ref{tab:grids}), with a spherical shell of
    35km of thickness, and density function with $b$ of 1km.
    The points inside the dashed line are the ones that present an
    error lower than 0.1\%.
    }
\label{fig:grid-search}
\end{figure}

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/exponential-delta.pdf}
\caption{
    Numerical error for different exponential density functions and values of the delta
    ratio.
    The top panels show the exponential density functions for different values of $b$
    (Eq. \ref{eq:density-exp}).
    The error computations were performed on the ``Big Grid'' configuration described in
    Table \ref{tab:grids}, with fixed values of the distance-size ratio
    $D$ obtained for the linear density case ($D=1, 2, 8$ for the
    potential, gradient components, and Marussi tensor components, respectively).
    The difference is reported as a percentage of the shell values.
    }
\label{fig:delta-exponential}
\end{figure*}

Having chosen values of $D$ equal to the linear density case, we are free to explore the
integration error as a function of $\delta$ in more detail and how it varies for
different values of $b$ (Eq. \ref{eq:density-exp}).
We perform the error calculations for the ``Big Grid'' configuration (Table
\ref{tab:grids}) for two spherical shell models with 1km and 35km thickness.
The calculations are repeated for different values of $b$ to examine the error variation
with the sharpness of the density function.
Because larger $\delta$ values result in fewer tesseroid divisions,
our intention is to find the highest value of $\delta$ whose numerical error is bellow
the $0.1\%$ threshold.

Fig. \ref{fig:delta-exponential} shows the density functions and the resulting relative
differences ($V$, $g_z$, and $g_{zz}$) for different values of $b$ and thickness of the
shell and tesseroid model.
The relative difference for $V$ and $g_z$ falls below the 0.1\% threshold for
$\delta = 0.2$ in both the thick and thin shell cases.
On the other hand, optimal values of $\delta$ for $g_{zz}$
are $\delta = 0.2$ for the thin shell and $\delta = 0.01$ for the thick shell.
Once again, we opt for the conservative value of $\delta = 0.01$ and sacrifice
performance for the sake of accuracy.

@@

We applied the new algorithms and optimal values of $D$ and $\delta$ determined
previously to calculate the gravitational effects of the Neuqu\'en Basin:
a sedimentary basin
located to the east of the Andes, between 32$^\circ$S and 40$^\circ$S latitude
(Fig. \ref{fig:neuquen-basin}a).
The basin includes continental and marine siliciclastics, carbonates, and evaporites
accumulated over the Jurassic and the Cretaceous constituting a stratigraphic record up
to 5000m of depth \citep{Howell2005}.

@@

The thickness of the sediment pack was digitized from \citet{Heine2007} on a regular
grid with a resolution of 0.05$^\circ$ on both longitude and latitude directions
(Fig. \ref{fig:neuquen-basin}b).
We created a tesseroid model of the sediment pack by placing a
$0.05^\circ \times 0.05^\circ$ tesseroid on each node of the grid.
The top of each tesseroid was fixed at 0 m depth and the bottom at corresponding
thickness of the basin.

We must also define a density function for the tesseroid model.
\citet{Sigismondi2012} measured a minimum and maximum density contrast for
the Neuqu\'en basin of -412kg/m$^3$ and -275kg/m$^3$, respectively.
We have chosen an exponential density variation (Eq. \ref{eq:density-exp}) that assumes
the minimum value on the top surface and the maximum at 5858m depth (the thickest part
of the basin), with a value of $b$ equal to 2km.
This density function can be seen on Fig. \ref{fig:neuquen-basin}c.

Finally, we computed the gravitational potential $V$, gradient components $g_x$,
$g_y$ and $g_z$, and the Marussi tensor components $g_{xx}$, $g_{xy}$,
$g_{xz}$, $g_{yy}$, and $g_{zz}$ on a computation grid of $159\times163$ nodes
($0.05^\circ$ spacing on both longitude and latitude) at a 50km height over the
reference ellipsoid.
The resulting fields can be seen in Fig.
\ref{fig:neuquen-basin}d-l.
We have excluded the other components of the Marussi tensor from the Fig.
\ref{fig:neuquen-basin},
although they can be found in the online repository.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]{figures/spherical-shell.pdf}
\caption{
    Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
    The computation point $Q$ is located on the $z$ axis at a distance $r$ from
    the origin of the coordinates system.
    For our purposes we will assume that $Q$ is outside of the shell,
    i.e. $r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}

Consider a spherical shell with inner radius $R_1$ and outer radius $R_2$,
whose density is a function $\rho(r')$ of the radial coordinate
(Fig. \ref{fig:spherical-shell}).
The gravitational potential generated by the shell on an arbitrary external
point $Q$ can be written as follows:

@@

\noindent where $\ell$ is defined in equation \ref{eq:ell}.

For $Q$ located along the $z$ axis (i.e., $\phi=90^\circ$) at a distance $r$ from the
origin, equation \ref{eq:ell} simplifies to:

@@

Because the computation point $Q$ is outside of the shell, $r > r'$ and the square roots
in equation \ref{eq:shell-pot-sqrts} simplify to

@@

From Eq. \ref{eq:shell-pot} we can obtain expressions of the gravitational potential for
different density functions.

>>>>

\title[Variable Density Tesseroids]{
    Gravitational field calculation in spherical coordinates using variable
    densities in depth
}
\author[S.R. Soler, A. Pesce, L. Uieda and M.E. Gimenez]{
    Santiago R. Soler$^{1,2}$, Agustina Pesce$^{1,2}$, Leonardo Uieda$^3$ and
    Mario E. Gimenez$^{1,2}$ \\
    $^1$CONICET, Argentina.~e-mail: santiago.r.soler@gmail.com\\
    $^2$Instituto Geofísico Sismológico Volponi, Universidad Nacional de
    San Juan, Argentina\\
    $^3$Department of Earth Sciences, SOEST, University of Hawai‘i at
    M\={a}noa, Honolulu, Hawaii, USA
}

@@

The lithosphere's density variation with depth has been studied for close to a
century.
Over this time period,
several density-depth relations have been proposed for different rock types
\citep[e.g.][]{Maxant1980, Rao1986, Rao1993, Rao1994}.
Furthermore, depth-variable densities have been used in forward and
inverse gravity modelling, mostly applied to sedimentary basins
\citep{Cordell1973, Rao1986, Cowie1990, Rao1993, Rao1994, Zhang2001,
Welford2010}.
These forward gravity models have been developed for two or three dimensional
bodies in Cartesian coordinates, which limits the applications to local scales.
The advent of satellite gravimetry has provided gravity field
measurements with global coverage, enabling modelling and interpretation on regional and
global scales.
Hence, designing forward modelling methods that reproduce the gravity anomalies for
such scales is of high importance.

@@

Two recent papers present other approaches for homogeneous tesseroids gravitational
fields computation and incorporate methodologies for variable density tesseroid with
specific density profiles.
\citet{Fukushima2018} analytically integrates the volumetric integral for the
gravitational potential in the radial direction, obtaining a surface integral, which is
then numerically solved by applying a conditionally splitting and the double exponential
quadrature rule. The gradient of the potential and the Marussi tensor components are
then computed by finite differences. \citet{Fukushima2018} also generalizes his method
to an arbitrary polynomial density profile on the radial coordinate.
\citet{Lin2018} made a comparison of the different methodologies for homogeneous
tesseroid gravitational fields computation by applying different discretizations
algorithms to each one of them.
From this analysis they develop a combined method:
for computation points near the tesseroid they apply a GLQ numerical integration with
a two dimensional adaptive discretization algorithm, i.e.~only in the horizontal
dimension. But, if the computation point is farther than a certain truncation distance,
then a second order Taylor series approximation is applied along with the regular
subdivision developed by \citet{Grombein2013}.
They also introduce a variation of their combined method to compute the gravitational
fields generated by tesseroids with linear density on the radial coordinate.

@@

Applying a $N$th order GLQ, we can approximate each integral in
equations~\ref{eq:tesseroid-pot} and~\ref{eq:tesseroid-grav} by a weighted sum of the
integration kernel evaluated on the roots of an $N$th order Legendre polynomial
\citep[p.~390]{Hildebrand1987}.
Unlike the homogeneous density case, the radial density function $\rho(r)$ must also be
included in the integration and evaluated on the Legendre polynomial roots
(i.e.~quadrature nodes).

@@

\begin{equation}
    \cos\psi_t =
        \sin\phi\sin\phi_t + \cos\phi\cos\phi_t\cos(\lambda - \lambda_t).
\end{equation}

\textit{Step 2}:
If none of the dimensions of the tesseroid fail inequality~\ref{eq:condition}, then
compute the gravitational effect of the tesseroid using a second-order GLQ
(Eq.~\ref{eq:glq-var-dens}).
Add the computed effect to a running total.

\textit{Step 3}:
If any dimension fails inequality~\ref{eq:condition}, split the tesseroid in half along
the offending dimensions.
Repeat steps 1-3 for all smaller tesseroids until none are left.

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]
    {figures/density-based-discretization-algorithm.pdf}
\caption{
    Example application of the density-based discretization algorithm to a non-linear
    density function.
    (a) the normalised density function $\rho_n(r')$ (blue), current boundaries of the
    tesseroid (orange dots), and the linear density function $\rho_l(r')$ (orange line).
    The dashed red line represents the maximum density difference $\Delta \rho (r')$ at
    which the tesseroid would be divided (assuming that the
    inequality~\ref{eq:delta-density} is not satisfied).
    (b) second iteration of the algorithm with a new linear density function and maximum
    density difference. The tesseroid would be divided at the depth indicated by the
    dashed red line.
    (c) third iteration of the algorithm.
    (d) final output of the density-based discretization, assuming that all four new
    tesseroids satisfy inequality~\ref{eq:delta-density}.
}
\label{fig:density-discretization-algorithm}
\end{figure*}

@@

The algorithm is comprised of the following steps
(Fig.~\ref{fig:density-discretization-algorithm}):

@@

\textit{Step 5}:
If inequality~\ref{eq:delta-density} is not satisfied, then the tesseroid is split in
two parts at the radius $r_\text{max}$ at which the maximum absolute difference
(Eq.~\ref{eq:density-abs-diff}) takes place.

@@

Once all smaller tesseroids satisfy Eq.~\ref{eq:delta-density},
each one is subjected to the two dimensional adaptive
discretization algorithm described earlier to calculate their gravitational effects.

@@

\citet{Uieda2016} compared the numerical integration of homogeneous density tesseroids
with the analytical solution of a spherical shell \citep{Mikuska2006, Grombein2013} in
order to obtain default values for the distance-size ratio $D$.
We will follow this idea but for our needs the spherical shell must
have the same density function of radius as our tesseroid model.
\citet{Lin2018} show the analytical solution of the gravitational potential generated by
a spherical shell with linear density in the radial coordinate.
We derive an expression for such gravitational potential in case the spherical shell has
an arbitrary density function in depth (Eq.~\ref{eq:shell-pot}), and obtained the
expressions in case of linear and exponential density (see Appendix~\ref{sec:shell}).

@@

\noindent
has an analytical solution given by Eq.~\ref{eq:shell-pot-linear}.

The absolute density difference defined on equation~\ref{eq:density-abs-diff} will
always be zero for the linear density case.
As a result, the inequality~\ref{eq:delta-density} will always be satisfied and no
divisions will ever be performed.
Therefore, the two dimensional adaptive discretization algorithm is the
only mechanism that controls the accuracy of the numerical integration.
For this reason, we will only determine the minimum value of $D$ needed in order to
guarantee an acceptable accuracy while ignoring the value of $\delta$.

In order to compare the numerical results with the analytical solution we
must build spherical shell models made of tesseroids.
These shell models are determined by their inner and outer radii and by the collection
of tesseroids used to build the shell.
For this analysis we have chosen to set the outer radii of every shell model equal to
the mean Earth radius (6378.137 km), while the inner radius is determined by the desired
thickness of the shell.
We choose  different values of shell thickness (100m, 1km, 10km, 100km and 1000km) in
order to account for different computation scenarios.
Furthermore, we propose different tesseroids collections to build the shell model by
regularly splitting it along the longitudinal and latitudinal directions.
Each collection, described in Table~\ref{tab:tess-collections}, is composed by
a different number of tesseroids, i.e.~with tesseroids of different sizes.
Finally, for each collection we build a shell model with the thicknesses
previously described, giving a total of 20 different shell models.

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/linear-density-diffs.pdf}
\caption{
    Differences between the gravitational fields generated by each tesseroid shell model
    and the analytical solution.
    Each model has a linear density function (Eq.~\ref{eq:density-linear}).
    The computations were performed on the four grids described in
    Table~\ref{tab:grids} and using 20 shell models built with the tesseroids
    collections detailed in Table~\ref{tab:tess-collections} and five different
    thicknesses.
    Each line represents the maximum absolute difference between every shell model with
    a certain thickness.
    Due to the linearity of the density function, the density-based discretization
    algorithm is not applied.
    Differences are reported as a percentage of the shell values.
    The horizontal dashed black line represents a target difference of 0.1\%.
}
\label{fig:D-linear}
\end{figure*}

@@

We observe from Fig.~\ref{fig:D-linear} that the relative errors for the
potential and the $g_z$ component fall below the 0.1\% threshold at
$D=1$ and $D=2.5$, respectively, for every shell model and computation grid.
These results also show that there's no meaningful dependency between the accuracy of
the method and the thickness of the shell, the collection of tesseroids used to build
it or the height of the computation grid (the results for the Satellite grid and the
Global grid are very similar).

@@

We compute the relative difference between the numerical and analytical results for the
gravitational potential ($V$) and the vertical component of its gradient ($g_z$) on the
computation grids detailed in Table~\ref{tab:grids} and using shell models with the same
thicknesses as the linear density case.
Because this is a time consuming computation, we limit the analysis only to the second
collection of tesseroids in Table~\ref{tab:tess-collections}.
For the sake of brevity, we compute the maximum differences between the results
involving each computation grid, shell thickness and value of the $b$ factor.
The results are shown in Fig.~\ref{fig:grid-search}, in which points inside the dotted
lines are the ones that present a numerical error lower than the 0.1\% threshold.
Also shown in Fig.~\ref{fig:grid-search} are the $D$ values determined for the linear
density function in the previous section ($D_\text{linear}$).
The values of $D_\text{linear}$ are within the 0.1\% threshold and are the optimum
values of $D$ for $V$ and $g_z$.
These results indicate that the values of $D_\text{linear}$ can be safely extrapolated
to the exponential case.

@@

\begin{figure}
\centering
\iftwocol{
\includegraphics[width=0.9\linewidth]
    {figures/grid-search.pdf}
}{
\includegraphics[width=0.5\linewidth]
    {figures/grid-search.pdf}
}
\caption{
    Numerical error exploration in the $D$-$\delta$ space.
    The percentage difference values were obtained from the comparison
    between the analytical solution and the numerical approximation of
    the gravitational fields ($V$, $g_z$ and $g_{zz}$) generated by a
    spherical shell with an exponential density function (Eq.~\ref{eq:density-exp}).
    These comparisons were carried out on the ``Big Grid'' (Table~\ref{tab:grids}), with
    a spherical shell of 35km of thickness, and density function with $b$ of 1km.
    The points inside the dashed line are the ones that present an
    error lower than 0.1\%.
    }
\label{fig:grid-search}
\end{figure}

@@

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/exponential-delta.pdf}
\caption{
    Numerical error for different exponential density functions and values of the delta
    ratio.
    The top panels show the exponential density functions for different values of $b$
    (Eq.~\ref{eq:density-exp}).
    The error computations were performed on the ``Big Grid'' configuration described in
    Table~\ref{tab:grids}, with fixed values of the distance-size ratio
    $D$ obtained for the linear density case ($D=1, 2, 8$ for the
    potential, gradient components, and Marussi tensor components, respectively).
    The difference is reported as a percentage of the shell values.
    }
\label{fig:delta-exponential}
\end{figure*}

Having chosen values of $D$ equal to the linear density case, we are free to explore the
integration error as a function of $\delta$ in more detail and how it varies for
different values of $b$ (Eq.~\ref{eq:density-exp}).
We perform the error calculations for the ``Big Grid'' configuration
(Table~\ref{tab:grids}) for two spherical shell models with 1km and 35km thickness.
The calculations are repeated for different values of $b$ to examine the error variation
with the sharpness of the density function.
Because larger $\delta$ values result in fewer tesseroid divisions,
our intention is to find the highest value of $\delta$ whose numerical error is bellow
the $0.1\%$ threshold.

Fig.~\ref{fig:delta-exponential} shows the density functions and the resulting relative
differences ($V$, $g_z$, and $g_{zz}$) for different values of $b$ and thickness of the
shell and tesseroid model.
The relative difference for $V$ and $g_z$ falls below the 0.1\% threshold for
$\delta = 0.2$ in both the thick and thin shell cases.
On the other hand, optimal values of $\delta$ for $g_{zz}$
are $\delta = 0.2$ for the thin shell and $\delta = 0.01$ for the thick shell.
Once again, we opt for the conservative value of $\delta = 0.01$ and sacrifice
performance for the sake of accuracy.

@@

We applied the new algorithms and optimal values of $D$ and $\delta$ determined
previously to calculate the gravitational effects of the Neuqu\'en Basin:
a sedimentary basin
located to the east of the Andes, between 32$^\circ$S and 40$^\circ$S latitude
(Fig.~\ref{fig:neuquen-basin}a).
The basin includes continental and marine siliciclastics, carbonates, and evaporites
accumulated over the Jurassic and the Cretaceous constituting a stratigraphic record up
to 5000m of depth \citep{Howell2005}.

@@

The thickness of the sediment pack was digitized from \citet{Heine2007} on a regular
grid with a resolution of 0.05$^\circ$ on both longitude and latitude directions
(Fig.~\ref{fig:neuquen-basin}b).
We created a tesseroid model of the sediment pack by placing a
$0.05^\circ \times 0.05^\circ$ tesseroid on each node of the grid.
The top of each tesseroid was fixed at 0 m depth and the bottom at corresponding
thickness of the basin.

We must also define a density function for the tesseroid model.
\citet{Sigismondi2012} measured a minimum and maximum density contrast for
the Neuqu\'en basin of -412kg/m$^3$ and -275kg/m$^3$, respectively.
We have chosen an exponential density variation (Eq.~\ref{eq:density-exp}) that assumes
the minimum value on the top surface and the maximum at 5858m depth (the thickest part
of the basin), with a value of $b$ equal to 2km.
This density function can be seen on Fig.~\ref{fig:neuquen-basin}c.

Finally, we computed the gravitational potential $V$, gradient components $g_x$,
$g_y$ and $g_z$, and the Marussi tensor components $g_{xx}$, $g_{xy}$,
$g_{xz}$, $g_{yy}$, and $g_{zz}$ on a computation grid of $159\times163$ nodes
($0.05^\circ$ spacing on both longitude and latitude) at a 50km height over the
reference ellipsoid.
The resulting fields can be seen in Fig.~\ref{fig:neuquen-basin}d-l.
We have excluded the other components of the Marussi tensor from the
Fig.~\ref{fig:neuquen-basin}, although they can be found in the online repository.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\linewidth]{figures/spherical-shell.pdf}
\caption{
    Spherical shell with inner and outer radii $R_1$ and $R_2$, respectively.
    The computation point $Q$ is located on the $z$ axis at a distance $r$ from
    the origin of the coordinates system.
    For our purposes we will assume that $Q$ is outside of the shell,
    i.e.~$r > R_2$.
}
\label{fig:spherical-shell}
\end{figure}

Consider a spherical shell with inner radius $R_1$ and outer radius $R_2$,
whose density is a function $\rho(r')$ of the radial coordinate
(Fig.~\ref{fig:spherical-shell}).
The gravitational potential generated by the shell on an arbitrary external
point $Q$ can be written as follows:

@@

\noindent where $\ell$ is defined in equation~\ref{eq:ell}.

For $Q$ located along the $z$ axis (i.e., $\phi=90^\circ$) at a distance $r$ from the
origin, equation~\ref{eq:ell} simplifies to:

@@

Because the computation point $Q$ is outside of the shell, $r > r'$ and the square roots
in equation~\ref{eq:shell-pot-sqrts} simplify to

@@

From Eq.~\ref{eq:shell-pot} we can obtain expressions of the gravitational potential for
different density functions.
